The Power of Two Random Choices\\ \large A follow up seminar to the Parallel Algorithms lecture\footnote{Jesper Larsson Träff, lecture ``Parallel Algorithms'', 2012 winter term at TU Wien.}

### 2014-01-24 article/article.tex
some review

Consider a variation of the above \emph{single-choice balls-into-bins} strategy: For each ball, chose $d \geq 2$ bins independently and uniformly at random and place the ball into the \emph{least full bin}\footnote{If there are multiple bins with the same minimal load, chose one among them arbitrarily.}. This simple \emph{multiple-choice balls-into-bins} scheme, which we will call the \emph{Greedy} algorithm, achieves a maximum load of $\frac{m}{n} + \Theta\left(\frac{\log \log n}{\log d}\right)$ w.h.p.~\cite{ABKU99}~\cite{BCSV06}. This apperantly minor change introduces a small amount of choice which, compared to the single-choice scheme, leads to a significant reduction in the maximum load. Curiously, the exponential decrease in the gap is already achieved having just two choices (that is, $d=2$). Each additional choice decreases the bound for the maximum gap by just a constant factor~\cite{MRS01}. Extensive research following the influental paper by Azar et. al \cite{ABKU99}, who were the first to explicitly prove this phenomenon, suggests that this \emph{two-choice paradigm} is not just an artificat of the simple balls-into-bins model, but a general principle applicable to a variety of problems.

The paper is organized as follows: We will first introduce different variations of the \emph{balls-into-bins problem} in Section \ref{sec:classification} to provide some overview of different directions in research as well as possibilities of abstraction from applications, some of which are listed in Section \ref{sec:applications}. Sections \ref{sec:single-choice} to \ref{sec:multiple-choice} will deal with the most basic variation of the balls-into-bins problem, namely the sequential, finite and static variant (see Section \ref{sec:classification} for details). For comparison and as a starting point, Section \ref{sec:single-choice} provides the results for the single-choice balls-into-bins problem. Section \ref{sec:multiple-choice} provides an in-depth presentation and analysis of the a basic \emph{multiple-choice balls-into-bins} scheme, the \emph{Greedy} algorithm.

Surprisingly, V\"ocking \cite{VOC03} showed that \emph{asymetric tie-breaking} in combination with \emph{partitioning} the $n$ bins into $d$ groups can further reduce the bound for the gap. Since the scheme is somewhat techniqual, we refer the reader to Section \ref{sec:AlwaysGoLeft} for details. \todo{more here?}

@@

In the \emph{single-choice balls-into-bins} strategy, each ball is placed into a bin choosen uniformly and independently at random.

@@

\section{Multiple-choice balls-into-bins}
\label{sec:multiple-choice}
The \emph{single-choice balls-into-bins strategy} (Section \ref{sec:single-choice}) chooses one bin for each ball. A \emph{multiple-choice balls-into-bins strategy} generalizes this idea by choosing $d$ bins and placing the ball into the least full bin. We will discuss two different multiple-choice schemes: The \emph{Greedy$[d]$} scheme (Section \ref{sec:greedy}) places the ball into the least full bin among $d$ bins chosen uniformly and independently at random whereas the \emph{Always-Go-Left} scheme (Section \ref{sec:AlwaysGoLeft}) uses a nonuniform selection of $d$ bins.

@@

\todo{analysis}

>>>>

Consider a variation of the above \emph{single-choice balls-into-bins} strategy: For each ball, chose $d \geq 2$ bins independently and uniformly at random and place the ball into the \emph{least full bin}\footnote{If there are multiple bins with the same minimal load, chose one among them arbitrarily.}. This simple \emph{multiple-choice balls-into-bins} scheme, which we will call the \emph{Greedy} algorithm, achieves a maximum load of $\frac{m}{n} + \Theta\left(\frac{\log \log n}{\log d}\right)$ w.h.p.~\cite{ABKU99}~\cite{BCSV06}. This apperantly minor change introduces a small amount of choice which, compared to the single-choice scheme, leads to a significant reduction in the maximum load. 
\begin{comment}
Curiously, the exponential decrease in the gap is already achieved having just two choices (that is, $d=2$). Each additional choice decreases the bound for the maximum gap by just a constant factor~\cite{MRS01}. 
\end{comment}
Extensive research following the influental paper by Azar et. al \cite{ABKU99}, who were the first to explicitly prove this phenomenon, suggests that this \emph{two-choice paradigm} is not just an artificat of the simple balls-into-bins model, but a general principle applicable to a variety of problems.

Condsider a slightly more involved strategy, which we will call the \emph{Always-go-left} scheme: Suppose that the $n$ bins are partitioned evenly into $d$ groups, each of size $\lfloor \frac{n}{d}\rfloor$ or $\lceil\frac{n}{d}\rceil$ \todo{simplify to n/d?}. We select one bin from each group uniformly and independently at random and put the ball into the least full of those $d$ bins. In case a tie occurs, put the ball into the \emph{left-most} bin. That is, assuming a fixed ordering of bins $b_i$ for $1 \leq i \leq n$, from among the $d' \leq d$ bins with same minimal load, we put the ball into the bin with smallest index $i$. Somewhat surprisingly, V\"ocking \cite{VOC03} showed that such a strategy employing \emph{asymetric tie-breaking} in combination with \emph{partitioning}\todo{check} reduces the bound for the gap to $\Theta \left( \frac{\log\log n}{d \log \Phi_d}\right)$ \todo{$\Phi_d$ is undefined}.

The paper is organized as follows: We will first introduce different variations of the \emph{balls-into-bins problem} in Section \ref{sec:classification} to provide some overview of different directions in research as well as possibilities of abstraction from applications, some of which are listed in Section \ref{sec:applications}. Sections \ref{sec:single-choice} to \ref{sec:multiple-choice} will deal with the most basic variation of the balls-into-bins problem, namely the sequential, finite and static variant (see Section \ref{sec:classification} for details). For comparison and as a starting point, Section \ref{sec:single-choice} provides the results for the single-choice balls-into-bins problem. \todo{TODO: multiple-choice, analysis, mention that witness tree method very helpful}

@@

In the \emph{single-choice balls-into-bins} strategy, each ball is placed into a bin choosen uniformly and independently at random. \todo{more specific?}

@@

\section{Multiple-choice balls-into-bins}
\label{sec:multiple-choice}
The \emph{single-choice balls-into-bins strategy} (Section \ref{sec:single-choice}) chooses one bin for each ball. A \emph{multiple-choice balls-into-bins strategy} generalizes this idea by choosing $d$ bins and placing the ball into the least full bin.

\todo{discuss the 3 classes}

We will discuss two different multiple-choice schemes: The \emph{Greedy$[d]$} scheme (Section \ref{sec:greedy}) places the ball into the least full bin among $d$ bins chosen uniformly and independently at random whereas the \emph{Always-Go-Left} scheme (Section \ref{sec:AlwaysGoLeft}) uses a nonuniform selection of $d$ bins.

@@
